---
title: "Correlation, network and heatmap plots"
subtitle: "Read and format raw data and perform statistical testing"
date: "`r Sys.Date()`"
output: html_document
header-includes: |
    \usepackage{caption}
    \usepackage{placeins}
    \usepackage{subfig}
---

## Load R packages and read in data

```{r message=FALSE}
library(tidyverse)
library(readxl)
library(stringr)
library(psych)
library(gdata)
library(qgraph)
library(kableExtra)
```


```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '/Users/xrydbh/GU/projects/donna_mega_cytokiner/methods')
```

```{r}
df.0.0  <- as.data.frame(read_excel("../file_for_publication/donna_mega_file_for_publication.1.1.xlsx", sheet = 1))
```

## This section can be read/edited after the first default output has been generated

> **_NOTE:_**  These 3 R markdown scripts can be used to generate plots based on log10 values of cytokine concentration measurements for time point b or based on log10 of the area under the curve of the concentration measurements for time interval c,d and e. The option is set by commenting out one of the two lines below. Make sure to put the output in a folder marked as time_point_b or time_interval_cde resepectivley before rerunning with new setting of time_span_of_interest.

```{r}
################## comment out one of the options below ##################
 #time_span_of_interest <- "timepoint_b"
time_span_of_interest <- "timeinterval_cde"
```


```{r}
df.0.0 <- df.0.0 %>% rename("GAD_continuos"=GAD)
if (time_span_of_interest=="timepoint_b"){
  #set name suffix and remove it from names of concentration variables
  names_suffix <- "_b_log10"  
  cyt_names <- df.0.0 %>% 
  names() %>% 
  str_subset(names_suffix)    # create list of cytokien names by removing suffix
  # need to remove ticde_ and ti_cde_ to avoid name dupliclation
  df.0.0 <- df.0.0  %>% select(-contains(c("ticde","ti_cde")))
  # remove dich_by_median_ from dichotomies variables
  df.0.0 <- df.0.0  %>%   rename_at(.vars = vars(starts_with("dich_by_median_")),
            .funs = funs(sub("dich_by_median_", "", .)))
  #remove tpb_ from GAD and brain volume variables
  df.0.0 <- df.0.0  %>%   rename_at(.vars = vars(starts_with("tpb_")),
            .funs = funs(sub("tpb_", "", .)))                            
}else if(time_span_of_interest=="timeinterval_cde"){
  # set and removes names suffix af AUC variables
  names_suffix <- "_log10_avg_cde_auc"
  cyt_names <- df.0.0 %>% 
  names() %>% 
  str_subset(names_suffix)    # create list of cytokien names by removing suffix
  # need to remove tpb_to avoid name dupliclation
  df.0.0 <- df.0.0  %>% select(-contains(c("tpb_")))
  # remove dich_by_median_ from dichotomies variables
  df.0.0 <- df.0.0  %>%   rename_at(.vars = vars(starts_with("dich_by_median_")),
            .funs = funs(sub("dich_by_median_", "", .)))
  #remove ticde_ from GAD variable
  df.0.0 <- df.0.0  %>%   rename_at(.vars = vars(starts_with("ticde_")),
            .funs = funs(sub("ticde_", "", .)))  
  #remove ti_cde_ from brain volume variables
  df.0.0 <- df.0.0  %>%   rename_at(.vars = vars(starts_with("ti_cde_")),
            .funs = funs(sub("ti_cde_", "", .)))                           
}else{
  print("you have to set the time_span_of_interest variable to either timepoint_b or timeinterval_cde")
}
```

Defining two vectors here one for all analyzed cytokines and one for all analyzed growth factors

```{r}
# make list of cytokine names and list of growth factor names
cyts0.0 <- c("IL1b","IL2","IL4","IL5","IL6","IL7","IL8","IL10","IL12p70","IL13","IL17","TNFa","MIP1b","MCP1MCAF","IFNg","GMCSF","GCSF")
gfs0.0 <- c("IGF1","BDNF","EPO","PDGF_BB","VEGF_A")
```


## Select (and rename) relevant columns
The relevant columns are  "ID", "Maternal_delivery_data","MAX_ROP","BPD", the dichotomised brain volume variables and "concentrations" (of cytokines and growth factors)


```{r}
clin_outs <- c("Maternal_delivery_data_Other0_Inflammation1_Hypoxia2","MAX_ROP","BPD","GAD","totalBrain","greyMatt","whiteMatt","whiteMatt","centStruct","cereb")

vars_to_select <- c("ID",clin_outs,cyt_names)
df.0.1 <- df.0.0  %>% 
  select(all_of(vars_to_select)) # select only columns that will be used
  
df.0.2  <- df.0.1 %>% 
  rename(matInfl = Maternal_delivery_data_Other0_Inflammation1_Hypoxia2) # shorten column name of maternal delivery variable

```

## Recode (and group) values
So that all samples are divided into 3 groups coded as 1,0 and NA meaning have-, have-not- feature in question and those with missing values.
```{r}
# shorten the matInfl entry in the vector used for selecting used columns
clin_outs <- str_replace(all_of(clin_outs), "Maternal_delivery_data_Other0_Inflammation1_Hypoxia2", "matInfl") 

#df.0.2
#df1$Sp2[df1$Sp2 == 8] <- 800
#recode(num_vec, `2` = 20L, `4` = 40L)
df.0.2.5 <- df.0.2 %>% 
mutate(MAX_ROP = recode(MAX_ROP, `1` = 0L, `2`=0L, `0`=0L, `3` = 1L),
       totalBrain = recode(totalBrain, `2` = 1, `1` = 0),
       greyMatt = recode(greyMatt, `2` = 1, `1` = 0),
       whiteMatt = recode(whiteMatt, `2` = 1, `1` = 0),
       centStruct= recode(centStruct, `2` = 1, `1` = 0),
       cereb = recode(cereb, `1` = 0, `2` = 1)
)
# df.0.2.5  %>% select(all_of(clin_outs))
```



## Collect clinical outcomes etc in clin_grouping and clin_grouping_val
Make data longer by collecting clinical outcomes.

### Alternative 1 make one grouped data frame with all groups have and havenot clinical outcome
```{r}
df.0.3 <- df.0.2.5 %>% pivot_longer(cols=all_of(clin_outs),names_to="clin_grouping",values_to="clin_grouping_val") %>%
  na.omit() %>% # remove samples with missing data
  group_by(clin_grouping,clin_grouping_val) # group by clin_grouping and value of that group
```
### Alternative 2 make grouped data frames, one for haves and one for havenots  clinical outcome
```{r}
# filter in clin_grouping_val==0 and group by clin_groupings
df.0.3_0 <- df.0.2.5 %>% pivot_longer(cols=all_of(clin_outs),names_to="clin_grouping",values_to="clin_grouping_val") %>%
na.omit() %>% 
filter(clin_grouping_val==0) %>% 
group_by(clin_grouping) 

# make list of lists of correlation values for clin_groupings, clin_grouping_val==1
df.0.3_1 <- df.0.2.5 %>% pivot_longer(cols=all_of(clin_outs),names_to="clin_grouping",values_to="clin_grouping_val") %>%
na.omit() %>% 
filter(clin_grouping_val==1) %>%  
group_by(clin_grouping)
```


```{r}
names(df.0.3) <- names(df.0.3) %>% str_replace(names_suffix,"") # remove information about time point and log from column names
group_counts <- df.0.3 %>% tally()  # look at sample counts for each value of maternal delivery variable

kbl(group_counts) %>%
  #kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
  kable_styling(bootstrap_options = "striped", full_width = F, position = "left")
```

```{r}
names(df.0.3_0) <- names(df.0.3_0) %>% str_replace(names_suffix,"") # remove information about time point and log from column names
group_counts_0 <- df.0.3_0 %>% tally()  # look at sample counts for each value of maternal delivery variable
names(df.0.3_1) <- names(df.0.3_1) %>% str_replace(names_suffix,"") # remove information about time point and log from column names
group_counts_1 <- df.0.3_1 %>% tally()  # look at sample counts for each value of maternal delivery variable
kbl(group_counts_0) %>%
#kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
kable_styling(bootstrap_options = "striped", full_width = F, position = "left")

kbl(group_counts_1) %>%
#kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
kable_styling(bootstrap_options = "striped", full_width = F, position = "left")
```

## Perform tests of pairwise correlation

Correlation tests are performed within the groups "inflammation" (n=) and "other" (n=) between each pairwise combination within the set of cytokines and growth factors. The results (r,p,n etc) are collected in a list with one entry for each group.

```{r}
corr_lst.0.0 <- df.0.3 %>% 
  group_split() %>% # split data frame by clinical grouping and clinical grouping variable
  map(select,-c("ID","clin_grouping","clin_grouping_val")) %>% # For each group select cytokines and growth factors
  map(psych::corr.test) # Create a list for each grouping variable with pairwise correlations and probability values between cytokines and growth factors 
```

### Give names to the list items of the list outputted by corr.test
```{r}
names(corr_lst.0.0) <- df.0.3 %>% group_keys() %>% unite(clin_groups,c("clin_grouping","clin_grouping_val")) %>% as.matrix()
```

## Rerun correlation test but now in separated runs to create separated lists of correlation results for clin_grouping_val==0 and clin_grouping_val==1 
This was done in order to be able to apply psych::paired.r with purrr:pmap and in order to easily plot triplets of correlation plots for grouping_val 0 and 1 and plots of difference in correlation for each grouping. 

```{r}
# make list of lists of correlation values for clin_groupings, clin_grouping_val==0
corr_lst_clin_group_0 <- df.0.3_0 %>% group_split() %>% 
map(select,-c("ID","clin_grouping","clin_grouping_val")) %>% 
map(psych::corr.test)

# Give names to the list items of the list outputted by corr.test
clin_grouping_vec <- df.0.3_0 %>% group_keys() %>% as.matrix()
clin_group_val_vec_0.0 <- rep("_0",length(clin_grouping_vec ))
clin_group_val_vec_0.1 <- paste(clin_grouping_vec,clin_group_val_vec_0.0,sep="")
names(corr_lst_clin_group_0) <- clin_group_val_vec_0.1 

corr_lst_clin_group_0.rval <- list() # make list of only rvals for clin_groupings, clin_grouping_val==0 
corr_lst_clin_group_0.pval <- list() # make list of only pvals for clin_groupings, clin_grouping_val==0
corr_lst_clin_group_0.n <- list()  # make list of only number of samples for clin_groupings, clin_grouping_val==0
for (i in 1:length(corr_lst_clin_group_0)){
  corr_lst_clin_group_0.rval[[i]] <- corr_lst_clin_group_0[[i]]$r # populate corr_lst_clin_group_0.rval list
  corr_lst_clin_group_0.pval[[i]] <- corr_lst_clin_group_0[[i]]$p
  corr_lst_clin_group_0.n[[i]] <- corr_lst_clin_group_0[[i]]$n
}

# make list of lists of correlation values for clin_groupings, clin_grouping_val==0 
corr_lst_clin_group_1 <- df.0.3_1 %>% group_split() %>% 
map(select,-c("ID","clin_grouping","clin_grouping_val")) %>% 
map(psych::corr.test)

# Give names to the list items of the list outputted by corr.test
clin_grouping_vec <- df.0.3_1 %>% group_keys() %>% as.matrix()
clin_group_val_vec_1.0 <- rep("_1",length(clin_grouping_vec ))
clin_group_val_vec_1.1 <- paste(clin_grouping_vec,clin_group_val_vec_1.0,sep="")
names(corr_lst_clin_group_1) <- clin_group_val_vec_1.1 

corr_lst_clin_group_1.rval <- list() # make list of only rvals for clin_groupings, clin_grouping_val==1 
corr_lst_clin_group_1.pval <- list() # make list of only pvals for clin_groupings, clin_grouping_val==1
corr_lst_clin_group_1.n <- list() # make list of only number of samples for clin_groupings, clin_grouping_val==1
for (i in 1:length(corr_lst_clin_group_1)){
  corr_lst_clin_group_1.rval[[i]] <- corr_lst_clin_group_1[[i]]$r # populate corr_lst_clin_group_1.rval list
  corr_lst_clin_group_1.pval[[i]] <- corr_lst_clin_group_1[[i]]$p
  corr_lst_clin_group_1.n[[i]] <- corr_lst_clin_group_1[[i]]$n
}

```

### Give names to the list items of the list outputted by corr.test
```{r}
names(corr_lst_clin_group_0.n) <- names(corr_lst_clin_group_0.pval) <- names(corr_lst_clin_group_0.rval) <- names(corr_lst_clin_group_0)
names(corr_lst_clin_group_1.n) <- names(corr_lst_clin_group_1.pval) <- names(corr_lst_clin_group_1.rval) <- names(corr_lst_clin_group_1)
```

## Perform tests of differences in pairwise correlation
Apply psych::paired.r with purrr:pmap 
```{r}

list.for.pair.r <- list(xy=corr_lst_clin_group_0.rval,xz=corr_lst_clin_group_1.rval,n=corr_lst_clin_group_0.n,n2=corr_lst_clin_group_1.n)
#group_by(clin_grouping) # group by clin_grouping and value of that group
#diffTest <- purrr::pmap(corr_lst_clin_group_0.rval,corr_lst_clin_group_1.rval,yz=NULL,corr_lst_clin_group_0.n, corr_lst_clin_group_1.n,psych::paired.r)# yz=NULL,n1, n2# Create a list for each grouping variable with pairwise correlations and probability values between cytokines and growth factors 
paired_corr.lst <- purrr::pmap(list.for.pair.r,psych::paired.r,yz=NULL)


paired_corr.lst.zval <- list()
paired_corr.lst.pval <- list()
for (i in 1:length(paired_corr.lst)){
  paired_corr.lst.zval[[i]] <- paired_corr.lst[[i]]$z # populate paired_corr.lst.zval list
  paired_corr.lst.pval[[i]] <- paired_corr.lst[[i]]$p
}

names(paired_corr.lst.zval) <- names(paired_corr.lst.pval) <- sub("_0","",names(corr_lst_clin_group_0.rval))
```

## Use loop to perform test instead
Seems to generate identicla results
```{r}
diff_in_corr <- list()
for (i in 1:length(corr_lst_clin_group_0.rval)) {
diff_in_corr[[i]] <- psych::paired.r(corr_lst_clin_group_0.rval[[i]],
                                     corr_lst_clin_group_1.rval[[i]],
                                     yz=NULL,
                                     corr_lst_clin_group_0.n[[i]], 
                                     corr_lst_clin_group_1.n[[i]])
}
```


## Save data objects for following steps
```{r}
if (time_span_of_interest=="timepoint_b"){
  fn ="001_read_and_format_data_and_stat_testing_tpb.RData"
}else if(time_span_of_interest=="timeinterval_cde"){
  fn ="001_read_and_format_data_and_stat_testing_ticde.RData"
}
```

```{r}
save(cyts0.0,
     gfs0.0,
     clin_outs,
     group_counts,
     group_counts_0,
     group_counts_1,
     corr_lst.0.0,
     corr_lst_clin_group_0.rval,
     corr_lst_clin_group_0.pval,
     corr_lst_clin_group_0.n,
     corr_lst_clin_group_1.rval,
     corr_lst_clin_group_1.pval,
     corr_lst_clin_group_1.n,
     paired_corr.lst.zval,
     paired_corr.lst.pval, 
     file = paste("./saved_data/",fn, sep=""))

```

```{r}
  save(time_span_of_interest,
  file = paste("./saved_data/time_span_of_interest.RData", sep=""))
```

