---
title: "Correlation, network and heatmap plots"
subtitle: "Format results of statistical testing"
date: "`r Sys.Date()`"
output: html_document
header-includes: |
    \usepackage{caption}
    \usepackage{placeins}
    \usepackage{subfig}
---

## Load R packages

```{r message=FALSE}
library(tidyverse)
library(readxl)
library(stringr)
library(psych)
library(gdata)
library(qgraph)
library(kableExtra)
```




```{r}
load(file = paste("./saved_data/time_span_of_interest.RData", sep=""))
if (time_span_of_interest=="timepoint_b"){
  fn ="001_read_and_format_data_and_stat_testing_tpb.RData"
}else if(time_span_of_interest=="timeinterval_cde"){
  fn ="001_read_and_format_data_and_stat_testing_ticde.RData"
}
```

## Load data from previous step
```{r}
loaded_objects_from_step_1 = load(file = paste("./saved_data/",fn, sep=""))
loaded_objects_from_step_1
```

## Filter r values of correlation matrix with values of p-value matrix

For the network plot we only want to display correlations that are significant.
Therefore we use the p-value matrix to filter the correlation matrix so that non significant correlation are set to zero.


```{r}

sign_corr_mats.lst <- list()

for (i in 1:length(corr_lst.0.0)){
  
  # Filter corelation matrix
  r.mat <- corr_lst.0.0[[i]]$r
  p.mat <- corr_lst.0.0[[i]]$p
  n.mat <- corr_lst.0.0[[i]]$n

  
  lowerTriangle(p.mat) <- upperTriangle(p.mat, byrow=TRUE)
  sign_corr.mat.0.0 = ifelse(p.mat < 0.05, r.mat , NA) # make all nonsinificant to na for oth
  sign_corr.mat.0.0[sign_corr.mat.0.0==1]<- NaN # also make r = 1 to NA since this is cyt vs cyt corr and dont need display
  sign_corr.mat.0.0 %>% as_tibble() %>% filter_all(any_vars(!is.na(.))) # remove samples that are all NA
  # reduce dfs so that every row and every column has at least one value
  dim(sign_corr.mat.0.0) # check dimensions before reduction
  sign_corr.mat.0.1  <- sign_corr.mat.0.0[,colSums(!is.na(sign_corr.mat.0.0 ))>0] # keep all columns with sum larger than 0
  sign_corr.mat.0.2  <- sign_corr.mat.0.1[rowSums(!is.na(sign_corr.mat.0.0 ))>0,] # keep all columns with sum larger than 0
  dim(sign_corr.mat.0.2) # check dimensions after reduction
  sign_corr.mat.0.2[is.na(sign_corr.mat.0.2)] <- 0 # replace NA with 0
  
  sign_corr_mats.lst[[i]] <- sign_corr.mat.0.2
}

```

## Make function that takes a list of correlation matrix and a list of pval matrixes and filter r values of correlation matrix with values of p-value matrix

For the network plot we only want to display correlations that are significant.
Therefore we use the p-value matrix to filter the correlation matrix so that non significant correlation are set to zero.


```{r}

reduce_cor_mat_to_only_sign <- function(r,p) {
  sign_corr_mats.lst <- list()

  for (i in 1:length(r)){
    
    # Filter corelation matrix
    r.mat <- r[[i]]
    p.mat <- p[[i]]
  
    lowerTriangle(p.mat) <- upperTriangle(p.mat, byrow=TRUE)
    sign_corr.mat.0.0 = ifelse(p.mat < 0.05, r.mat , NA) # make all nonsinificant to na for oth
    sign_corr.mat.0.0[sign_corr.mat.0.0==1]<- NaN # also make r = 1 to NA since this is cyt vs cyt corr and dont need display
    sign_corr.mat.0.0 %>% as_tibble() %>% filter_all(any_vars(!is.na(.))) # remove samples that are all NA
    # reduce dfs so that every row and every column has at least one value
    dim(sign_corr.mat.0.0) # check dimensions before reduction
    sign_corr.mat.0.1  <- sign_corr.mat.0.0[,colSums(!is.na(sign_corr.mat.0.0 ))>0] # keep all columns with sum larger than 0
    sign_corr.mat.0.2  <- sign_corr.mat.0.1[rowSums(!is.na(sign_corr.mat.0.0 ))>0,] # keep all columns with sum larger than 0
    dim(sign_corr.mat.0.2) # check dimensions after reduction
    sign_corr.mat.0.2[is.na(sign_corr.mat.0.2)] <- 0 # replace NA with 0
    
    sign_corr_mats.lst[[i]] <- sign_corr.mat.0.2
  }
sign_corr_mats.lst
}


```


## Apply function on corr_lst_clin_group_0 and corr_lst_clin_group_1
```{r}
sign_corr_lst_clin_group_0 <- reduce_cor_mat_to_only_sign(corr_lst_clin_group_0.rval,corr_lst_clin_group_0.pval)
sign_corr_lst_clin_group_1 <- reduce_cor_mat_to_only_sign(corr_lst_clin_group_1.rval,corr_lst_clin_group_1.pval)
```


## Create a list of significant cytokine names and growth factor names
The list will be used for coloring the nodes in the network plot respectively. The purpose of this chunk is to split the names vector of cytokines and growth factor into two list items (on for cytokines and one for growth factors). This will be used for color coding the nodes in the network plots.

```{r}

sign_cyts_and_gfs.lst <- list()

for (i in 1:length(corr_lst.0.0)){

  # select cytokine names
  cyts.oth.0.1 <- cyts0.0[cyts0.0 %in% row.names(sign_corr_mats.lst[[i]])] # check what cytokines that have significant correlation
  cyts.infl.0.1 <- cyts0.0[cyts0.0 %in% row.names(sign_corr_mats.lst[[i]])] # also for infl
  
  cyts.oth.0.2 <- seq(length(cyts.oth.0.1 )) # make incremental sequence from 1 to length of cyts that are significant
  
  # select growth factors other
  if(any(gfs0.0 %in% row.names(sign_corr_mats.lst[[i]]))){
    gfs.oth.0.1 <- gfs0.0[gfs0.0 %in% row.names(sign_corr_mats.lst[[i]])] # check what growth factors that have significant correlation
    gfs.oth.0.2 <- seq(max(cyts.oth.0.2)+1,length(c(cyts.oth.0.1,gfs.oth.0.1))) # make incremental sequence that starts where cytokine list  ended and incements with length of grothfactors that are significant
    gf_cyt_groups_corr.mat <- list(cytokines=cyts.oth.0.2,growthfactors=gfs.oth.0.2 )
  } else {
    gf_cyt_groups_corr.mat <- list(cytokines=cyts.oth.0.2)
  }
  sign_cyts_and_gfs.lst[[i]] <-  gf_cyt_groups_corr.mat
}
```

## Create a function that creates a list of significant cytokine names and growthfactor names
The list will be used for coloring the nodes in the network plot respectively. The purpose of this chunk is to split the names vector of cytokines and growth factor into two list items (on for cytokines and one for growth factors). This will be used for color coding the nodes in the network plots.

```{r}
get_names_of_significant_cyts_and_gfs <- function(corr_lst_clin_group.rval,sign_corr_mats.lst) {
  sign_cyts_and_gfs.lst <- list()
  
  for (i in 1:length(corr_lst_clin_group.rval)){#sign_corr_mats.lst
  
    # select cytokine names
    cyts.oth.0.1 <- cyts0.0[cyts0.0 %in% row.names(sign_corr_mats.lst[[i]])] # check what cytokines that have significant correlation
    cyts.infl.0.1 <- cyts0.0[cyts0.0 %in% row.names(sign_corr_mats.lst[[i]])] # also for infl
    
    cyts.oth.0.2 <- seq(length(cyts.oth.0.1 )) # make incremental sequence from 1 to length of cyts that are significant
    
    # select growth factors other
    if(any(gfs0.0 %in% row.names(sign_corr_mats.lst[[i]]))){
      gfs.oth.0.1 <- gfs0.0[gfs0.0 %in% row.names(sign_corr_mats.lst[[i]])] # check what growth factors that have significant correlation
      gfs.oth.0.2 <- seq(max(cyts.oth.0.2)+1,length(c(cyts.oth.0.1,gfs.oth.0.1))) # make incremental sequence that starts where cytokine list  ended and incements with length of grothfactors that are significant
      gf_cyt_groups_corr.mat <- list(cytokines=cyts.oth.0.2,growthfactors=gfs.oth.0.2 )
    } else {
      gf_cyt_groups_corr.mat <- list(cytokines=cyts.oth.0.2)
    }
    sign_cyts_and_gfs.lst[[i]] <-  gf_cyt_groups_corr.mat
  }
  sign_cyts_and_gfs.lst
}


```

## Apply get_names_of_significant_cyts_and_gfs
```{r}
sign_cyts_and_gfs.lst_0 <- get_names_of_significant_cyts_and_gfs(corr_lst_clin_group_0.rval,sign_corr_lst_clin_group_0)
sign_cyts_and_gfs.lst_1 <- get_names_of_significant_cyts_and_gfs(corr_lst_clin_group_1.rval,sign_corr_lst_clin_group_1)
```


## Filter z values of matrix of difference in correlation with values of corrsponding p-value matrix

We want the correlation plots of the difference in correlation to show only significant differences therefor we will filter out the non significant z-values from the matrix that was received as output.
Therefore we use the p-value matrix to filter the correlation matrix so that non significant correlation are set to zero.
(We would like the pairwise correlations between cytokines and/or growthfactors determine the thickness fo the edges.)

```{r}
cor_diff.lst <- list()
cor_diff_groups.lst  <- list()
for (i in 1:length(paired_corr.lst.zval)){
  z_score <- paired_corr.lst.zval[[i]]  
  p_val <- paired_corr.lst.pval[[i]]  
  
  # make function to replace NaN in df with 0
  is.nan.data.frame <- function(x)
  do.call(cbind, lapply(x, is.nan))
  # make copy of p_val
  p_val.0.0 <- p_val
  #apply function createdd above
  p_val.0.0 [is.nan(p_val.0.0 )] <- 1
  
  z_score.0.1 <- z_score
  # replace any z-values hta are zero  with 1. Why ????
  z_score.0.1[z_score.0.1==0] <- 1
  
  # set all unsignificant z-scores to NA
  sign_z = ifelse(p_val.0.0 < 0.05, z_score.0.1, NA)
  dim(sign_z)
  #Keep only columns with sum larger that 0
  sig_ele0<-sign_z[,colSums(!is.na(sign_z))>0] 
  dim(sig_ele0)
  #Keep only rows with sum larger that 0
  sign_z1<-sig_ele0[rowSums(!is.na(sig_ele0))>0,] 
  dim(sign_z1)
  #replace any na value with 0
  sign_z1[is.na(sign_z1)] <- 0
  
  
  # makes list of groups
  #gfs0.0 <- c("IGF1","BDNF","EPO","PDGF_BB","VEGF_A")
  #cyts0.0 <- c("IL1b","IL2","IL4","IL5","IL6","IL7","IL8","IL10","IL12p70","IL13","2IL172","TNFa","MIP1b","MCP1MCAF","IFNg","GMCSF","GCSF")
  cyts0.1<- cyts0.0[cyts0.0 %in% row.names(sign_z1)]
  cyts0.2 <- seq(length(cyts0.1))
  
  # 
  # gfs0.1 <- gfs0.0[gfs0.0 %in% row.names(sign_z1)]
  # gfs0.2 <- seq(max(cyts0.2)+1,length(c(cyts0.1,gfs0.1)))
  # gf_cyt_groups_corr_change <- list(cytokines=cyts0.2,growthfactors=gfs0.2)
  # 
  # 
  
  
   # select growth factors other
    if(any(gfs0.0 %in% row.names(sign_z1))){
      gfs0.1 <- gfs0.0[gfs0.0 %in% row.names(sign_z1)] # check what growth factors that have significant correlation
      gfs0.2 <- seq(max(cyts0.2)+1,length(c(cyts0.1,gfs0.1))) # make incremental sequence that starts where cytokine list  ended and incements with length of grothfactors that are significant
      gf_cyt_groups_corr_change <- list(cytokines=cyts0.2,growthfactors=gfs0.2)
    } else {
      gf_cyt_groups_corr_change <- list(cytokines=cyts0.2)
    }
  
  cor_diff.lst[[i]] <- sign_z1
  cor_diff_groups.lst[[i]] <- gf_cyt_groups_corr_change
}

names(cor_diff.lst) <- names(clin_outs)
```

## Prepare vectors with cytokines and growth factors with significant diffenrences in correlation  for each clinical group
For each clinical group, collect all cytokines and growth factors with significant changes in correlation into a vector. Create similar vectors with corresponding sums of z-values. Sort the vectors by the sums of z-values. bind vectors together as columns in a matrix

```{r}
#http://www.sthda.com/english/wiki/correlation-matrix-formatting-and-visualization
# ++++++++++++++++++++++++++++
# flattenCorrMatrix or in this case zMatrix
# ++++++++++++++++++++++++++++
# cormat : matrix of the correlation coefficients
# pmat : matrix of the correlation p-values
flattenCorrMatrix <- function(cormat, pmat) {
  ut <- upper.tri(cormat) # Returns a matrix of logicals the same size of a given matrix with entries TRUE in the lower or upper triangle.
  data.frame(
    row = rownames(cormat)[row(cormat)[ut]], # make vector of row names for each value of upper triangle ant set as first column
    column = rownames(cormat)[col(cormat)[ut]], # make vector of column names for each value of upper triangle ant set as first column
    cor  =(cormat)[ut], # make vector of each r-score or z-score  of upper triangle
    p = pmat[ut] # make vector of each p-value of upper triangle
  )
}
```


```{r}
ranked_peps <- list()
ranked_z <- list()
list_versions <- list()
for (i in 1:length(paired_corr.lst.zval)){
  Z <- paired_corr.lst.zval[[i]]  # put z-scores in array A
  normalized = (Z-min(Z,na.rm=T))/(max(Z,na.rm=T)-min(Z,na.rm=T))  #Normalized/standardize z-scores (make z scores vary between 1 and 0: Not really needed)
  B <- paired_corr.lst.pval[[i]] #put p-values in array B
  sign_z = ifelse(B < 0.05, normalized, NA)# filter in only significant z-scores
  
  #make tibbles
  tb.p <- as_tibble(B)
  tb.z <- as_tibble(sign_z)
  
  df.sign <- data.frame((tb.p < 0.05) * 1) # recode p-vals to binary
  m.sign <- as.matrix(df.sign) # make binary significance df into matrix again
  
  rownames(m.sign) <- colnames(m.sign) #assign rownames# renamed rownames(m.sign.z) to m.sign


  # make longer make tibble() and group by column
  # The matrices of z-vals and p-vals are made into a tibble with four columns row,col, cor and p-val.
  z_score_and_sig_long <- flattenCorrMatrix(sign_z, m.sign) %>% tibble() 
  
  # !!!!!!This step really dont seem necessary it is using p-val to filter in sing z-score as has allready been done!!!!!
  sig_z_score_long <- z_score_and_sig_long %>% mutate(sig_z=cor*p) # create new variable sig_z

  sig_z_score_long_row <- sig_z_score_long %>% group_by(row) # group by row (of upper triangle of correlation matrix)
  sig_z_score_long_col <- sig_z_score_long %>% group_by(column) # and group by column (of upper triangle of correlation matrix that has now been made longer)
  
  # sum z-scores per row/peptide and sort
  sum.z.sort_row <- sig_z_score_long_row %>% summarize(sum=sum(sig_z, na.rm = TRUE)) 
  # get number of sign z per row
  numb.z.row <- sig_z_score_long_row %>% na.omit() %>% tally()#summarize(n=n(sig_z, na.rm = TRUE)) 
  # sum z-scores per col/peptide and sort
  sum.z.sort_col <- sig_z_score_long_col %>% summarize(sum=sum(sig_z, na.rm = TRUE)) 
  # get number of sign z per col
  numb.z.col <- sig_z_score_long_col %>% na.omit() %>% tally()
 
  #join row and col sums
  sum.z.sort0.0 <- full_join(sum.z.sort_row, sum.z.sort_col, by = c("row" = "column")) 
  # join row and col counts of sign z (need this as an effect of making selecting upper triangle of correlation and making long)
  numb.z.sort0.0 <- full_join(numb.z.row, numb.z.col, by = c("row" = "column")) 
  # Since IL1b becomes NA in one column and VEGF_A in the other they need to be given value 0 so they dont crate probs later
  sum.z.sort0.1 <- sum.z.sort0.0 %>% 
    tidyr::replace_na(list(row=0,sum.x=0,sum.y=0)) 
  # Replace NA with zeros to avoid problems later
  numb.z.sort0.1 <- numb.z.sort0.0 %>% 
    tidyr::replace_na(list(row=0,n.x=0,n.y=0)) 
  
  #add row and col sums together, select, sort
  sum.z.sort <- sum.z.sort0.1 %>% mutate(sum_sig_z=sum.x+sum.y) %>% select(row,sum_sig_z) %>% arrange(by=desc(sum_sig_z))
  numb.z.sort <- numb.z.sort0.1 %>% mutate(numb_sig_z=n.x+n.y) %>% select(row,numb_sig_z) %>% arrange(by=desc(numb_sig_z))
  # sort numb.z.sort rows by sum.z.sort$row and select the row with counts
  counts_sig_z <- numb.z.sort[match(sum.z.sort$row, numb.z.sort$row),][,2]
  # bind cols to be pasted together
  peps_sum.z.sort.0 <- cbind(sum.z.sort[,1],counts_sig_z)
  # paste/unite sorted numb.z.sort$numb_sig_z to sum.z.sort$row
  peps_sum.z.sort.1 <- unite(as_tibble(peps_sum.z.sort.0), newCol)
  sum.z.sort[,1] <- peps_sum.z.sort.1 
  
  # rename row (previosly peptide) column to sub_clin_out variable value
  names(sum.z.sort)[names(sum.z.sort ) == "row"] <- names(paired_corr.lst.zval)[i]
  # rename sum_z column to sub_clin_out variable value
  names(sum.z.sort )[names(sum.z.sort ) == "sum_sig_z"] <- names(paired_corr.lst.zval)[i]
  # collect ranked peptides for sub_clin_out into ranked_peps
  ranked_peps[[i]] <- sum.z.sort[,1]
  # collect sums  of z values  for sub_clin_out inte ranked_z
  ranked_z[[i]] <- sum.z.sort[,2]

}

ranked_peps_tb <-bind_cols(ranked_peps) #table with only peptides
ranked_z_tb <-bind_cols(ranked_z) #table with only z-scores

ranked_peps_tb %>% names() #check the names of the tables
ranked_z_tb  %>% names()

#make z-val table long, group by clin_ut, summaraize and arrange by summed z-score
ranked_z_tb.sum.sort <- ranked_z_tb %>% pivot_longer(cols=names(ranked_z_tb),names_to = "clin_out",values_to = "z_score") %>% group_by(clin_out) %>% summarise(z_score = sum(z_score)) %>% arrange(by=desc(z_score))

ranked_peps_tb.0.0 <- ranked_peps_tb[ranked_z_tb.sum.sort$clin_out] # use clin_out col to order 
ranked_z_tb.0.0 <- ranked_z_tb[ranked_z_tb.sum.sort$clin_out]
```

## Save data objects for following steps

```{r}
if (time_span_of_interest=="timepoint_b"){
  fn ="002_format_results_of_statistical_testing_tpb.RData"
}else if(time_span_of_interest=="timeinterval_cde"){
  fn ="002_format_results_of_statistical_testing_ticde.RData"
}
```

```{r}
save(corr_lst.0.0,
     gfs0.0,
     sign_corr_mats.lst,
     sign_corr_lst_clin_group_0,
     sign_corr_lst_clin_group_1,
     sign_cyts_and_gfs.lst,
     sign_cyts_and_gfs.lst_0,
     sign_cyts_and_gfs.lst_1,
     cor_diff.lst,
     cor_diff_groups.lst,
     ranked_peps_tb.0.0,
     ranked_z_tb.0.0,
     file = paste("./saved_data/",fn, sep=""))

```
